---
title: "genre_format_data"
author: "Luke Novak"
date: "4/17/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analyzing the genre and format data:

Basic setup:
```{r setup}
library(dplyr)
library(ggplot2)
library(RPostgreSQL)
library(stringr)
require(DBI)

drv <- dbDriver("PostgreSQL")
db <- dbConnect(drv, user=user, host=host,
                password=pw, dbname='music_analysis') #DB DATA GOES HERE
```

To analyze the genre and format data we gathered through discogs (and the
python script scraper, we first need to grab our data from the database
and clean it up a little (although the data has already been cleaned)

To do this, we join the formats graph with the albums graph so we can see what
albums were released in what format
```{r get data}
formats <- dbGetQuery(db, "SELECT * FROM releases;")
all_albums <- dbGetQuery(db, "SELECT * FROM albums;")
formats_joined <- inner_join(formats, all_albums, by = 'id')
```

### Format analysis
To analyze the format of graphs over time (this was one of the main goals of 
the project) we take the formats of albums and the year in which they were 
released and group them together, summarizing them to get the count of each
format per year. We then proceed to remove outliers, and do a simple imputation
of data for the rest of 2019 to normalize it. Finally, we plot the data.
```{r format analysis}
count_format_per_year <- formats_joined %>% select(format, year) %>% 
  group_by(format, year) %>% summarize(total = n())

## We need to remove some outliers, like CDs prior to their inventin in 1982
count_format_per_year <- count_format_per_year %>% 
  filter(format != 'CD' | year > 1981)
## multiply the current year by 3 to extrapolate out
count_format_per_year <- count_format_per_year %>% 
  mutate(total = ifelse(year == 2019, total * 4, total))

ggplot(count_format_per_year, aes(x = year, y=total, color=format)) + geom_line()

```
As we can see, the format of albums has changed greatly over time. While Vinyl
is the only true contender for the beginning of the graph, we see the rapid
rise in CDs as time goes on, and the eventual downfall of CDs as well as the
vinyl revival in the mid 2000s. Finally we see the rise of digitial in the
2010s as well.

It is worth noting that Discogs is a biased site, and is thus heavily skewed
towards physical releases (vinyl especially), and so we must consider that when 
tagging the primary release of an album, users will try to refrain from using
the file tag unless no physical release is available. Thus, the file data is
much lower than expected

### Genre analyisis
To analyze genres, we tidy the data in a similar manner to the format tidying,
dropping all other columns, summarizing after a grouping with year, and
imputing for the rest of 2019. Since rock was the dominant genre, I also take
a look at non-rock genres, and then remove the "small genres" at the bottom
of the graph to make it easier to read. This can be treated like outlier
removal
```{r genres}
genres <- all_albums %>% filter(genre != 'None')
count_genre_per_year <- genres %>% select(genre, year) %>% 
  group_by(genre, year) %>% summarize(total = n())

count_genre_per_year <- count_genre_per_year %>% 
  mutate(total = ifelse(year == 2019, total * 4, total)) %>% filter(year > 1955)


ggplot(count_genre_per_year, aes(x = year, y=total, color=genre)) + geom_line()

count_genre_no_rock <- count_genre_per_year %>% filter(genre != 'Rock' & year > 1960)

ggplot(count_genre_no_rock, aes(x = year, y=total, color=genre)) + geom_line()

count_genre_just_big_ones <- count_genre_no_rock %>% 
  filter (genre != 'Non-Music' & genre != 'Children\'s' & genre != 'Classical' &
          genre != 'Stage & Screen' & genre != 'Reggae' & genre != 'Blues' & year != 2019)

ggplot(count_genre_just_big_ones, aes(x = year, y=total, color=genre)) + geom_line()
```

The genre graph was by far one of the more interesting that we had, as it shows
a vast number of cultural trends. There are too many to name here, but a few
examples to observe is the rise in hip-hop since the 80s (and ensuing fall of
rock in the 2010s that industry experts have been parroting for years), the
two Electronic booms, one in the 80s (sythwave), and a massive boom/bust in
the 2010s (the EDM Boom). Also observe the rise and fall and rise again of
Funk/Soul in the 70s, then returning in the 2010s.