---
title: "Tempo Analysis"
author: "Jessica Cheng"
date: "April 17, 2019"
output: html_document
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries and connect to database

```{r}
library(dplyr)
library(spotifyr)
library(RPostgreSQL)
library(tidyverse)
library(ggplot2)
require(DBI)

# loads the PostgreSQL driver
drv <- dbDriver("PostgreSQL")
# creates a connection to the postgres database
db <- dbConnect(drv, dbname = "music_analysis",
                 host = "fernando.nluken.com", port = 5432,
                 user = username, password = pw) #DB DATA GOES HERE

```

# Song Data
We joined tables songs and albums on album ID in our query to obtain information across both tables such as tempo and duration from songs table and year from albums table since they are the main features we were interested in studying. We queried all song data between 1960 and 2019 as this was the time period we were studying and ordered the data by year.

```{r}
# all tempo data between 1960 and 2019
df_tempo_all <- dbGetQuery(db, "SELECT tempo, duration, year 
                           FROM albums INNER JOIN songs ON albums.id = songs.album_id 
                           WHERE year BETWEEN 1960 AND 2019 
                           ORDER BY year")
```

# Identify outliers in song data and remove them
We want to study how tempo has changed over the years. First, we decided to create a boxplot to visualize any tempo outliers from each year. 

## Graph tempo by year and visualize outliers in song data
```{r}
# visualize outliers
tempo_all_plot <- ggplot(df_tempo_all, 
                         mapping = aes(x = as.factor(year), 
                                       y = tempo, 
                                       fill= as.factor(year))) + 
  geom_boxplot() + 
  labs(x = "Year", y = "Tempo") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position="none")
tempo_all_plot
```
Based on the boxplot produced, we see that there are some tempo outliers. Instead of fully removing these outlier as that would produce uneven column lengths, we decide to reassign each outlier value with NA. A new dataframe was produced which contained NA values of the tempo outliers. Then, the rows containing the NAs or outliers were filtered out so that the dataset would not contain outliers.
```{r}
# function to remove outliers (outside of 1.5*IQR range) and replace them with NAs
remove_outliers <- function(x, na.rm = TRUE) {
  qnt <- quantile(x, probs=c(.25, .75), na.rm = na.rm)
  H <- 1.5 * IQR(x, na.rm = na.rm)
  y <- x
  y[x < (qnt[1] - H)] <- NA
  y[x > (qnt[2] + H)] <- NA
  y
}

#info on df_tempo_all
str(df_tempo_all)

# distinct years
years <- df_tempo_all %>% distinct(year)
year_list <- years$year

# function to produce table for given year
table_year <- function(year_num) {
  df_tempo_all %>% filter(year == year_num)
}

# list of tables for each distinct year
list_tables <- lapply(year_list, table_year)

#function that takes in a df and produces a vector of tempo from df
year_tempo <- function(ydf) {
  ydf$tempo
}

# list of list of tempos for each year's tempo table
list_year_tempo <- lapply(list_tables, year_tempo)

# apply replace outliers with NAs to each list of list of tempos from each year
list_no_outliers <- lapply(list_year_tempo, remove_outliers)

# combines lists of tempos into one list
tempo_no_outliers <- unlist(list_no_outliers)
# check length of tempos list
length(tempo_no_outliers) # 246778
# extracts years from original tempo df
tempo_years <- df_tempo_all$year
# make sure length of years same as length of tempos list
length(tempo_years) #246778

# create df of tempo data
df <- cbind(tempo_no_outliers, tempo_years)
tempo_table <- as.data.frame(df)
names(tempo_table) <- c("tempo", "year")
tempo_table

#count how many outliers as NAs
na_tempo_count <- tempo_table %>% 
  summarise(na_count = sum(is.na(tempo)))
# tempo df with outlier/NAs removed
tempo_df <- tempo_table %>% filter(!is.na(tempo))
#check na rows have been removed
na_tempo_new_count <- tempo_df %>% 
  summarise(na_count = sum(is.na(tempo)))
```
## Visualize boxplot of new song dataframe with outliers removed

The new dataframe still contains a few outliers but as the size of the data is large enough, it was determined that they should not have a large effect on the mean tempo for each year.
```{r}
# new boxplot with NAs outliers removed
tempo_df_plot <- ggplot(tempo_df, 
                        mapping = aes(x = as.factor(year), 
                                      y = tempo, 
                                      fill= as.factor(year))) + 
  geom_boxplot() + 
  labs(x = "Year", y = "Tempo") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position="none")
tempo_df_plot
```
# Analysis of tempo

## Analyze relationship between tempo over year

### Graph of average tempos over year
As our question involves studying how tempo has changed over time, we decided to determine the average tempo for each year. Based on observation alone, the average tempo of songs seems to increase over time from 1960 until 1980. Between 1980 and 1999 the average tempo of the songs decreased. Then, the average tempo seems to generally increase again by 2000. The lowest tempos were around 1960s and around 1980 and recent years, the average tempos were quite high.

```{r}
#Average tempos
tempo_avg <- tempo_df %>% group_by(year) %>% 
  summarise(avg_tempo = mean(tempo))
# plot avg tempo over time with outliers/ NAs removed
avg_tempo_plot <- ggplot(tempo_avg, 
                         mapping = aes(x = year, 
                                       y = avg_tempo, 
                                       color = avg_tempo)) + 
  geom_point() + 
  labs(x = "Year", y = "Average Tempo") + 
  ggtitle("Average Tempo Over Time") +
  theme(plot.title = element_text(hjust = 0.5))
avg_tempo_plot + scale_color_gradient()
```
### Analyze correlation and model of tempo and year
Correlation is a statistical measure that suggests the level of linear dependence between two variables. The correlation between the average tempo and year was 0.65. There is a moderately positive correlation between average tempo and year. In general, as year increased since 1960, so did tempo. However, between 1980 and 1990, there was a decrease in tempo which lowered the positive correlation between average tempo and year. One of the other factors we want to study is genre. Certain genres have a specific range of tempos. The popularity of certain genres per year may explain why the average tempo changed per year.

Now that we have seen the linear relationship pictorially in the scatter plot and by computing the correlation, we decide to build the linear model. We have established the relationship between the predictor and response in the form of a mathematical formula for tempo as a function for duration. To ensure that the regression model is statistically significant, we observe the p-value from summary statistics. We can consider a linear model to be statistically significant only when both these p-Values are less that the pre-determined statistical significance level, which is ideally 0.05. This is visually interpreted by the significance stars at the end of the row. The more the stars beside the variable's p-Value, the more significant the variable. The p-value from this model was 2.03e-08 so year is significant.

When there is a p-value, there is a hull and alternative hypothesis associated with it. In Linear Regression, the Null Hypothesis is that the coefficients associated with the variables is equal to zero. The alternate hypothesis is that the coefficients are not equal to zero. There exists a relationship between the independent variable in question (year in this case) and the dependent variable (tempo). As the p-value is below 0.05, this means that the null hypothesis is rejected and there appears to be a relationship between year and average tempo.

What R-Squared tells us is the proportion of variation in the dependent (response) variable that has been explained by this model. The R-squared is 0.42. About 40% of the variation in the average tempo has been explained by this model.
```{r}
# correlation
cor(tempo_avg$avg_tempo, tempo_avg$year) # around 0.65

#lm
tempo_mod <- lm(avg_tempo ~ year, data = tempo_avg)
summary(tempo_mod)
```
## Analyze relationship between tempo and duration

### Scatterplot of duration vs tempo
It was hypothesized that the duration of songs may also have an effect on tempo. For instance, a longer song may have slower tempo. We decided to create a scatterplot to observe if there is a linear relationship. There did not seem to be a linear relationship. What was interesting was that the range of tempos did seem to decrease over duration based on observation alone. However, further analysis was needed.

```{r}
# duration vs tempo
dur_tempo_plot <- ggplot(df_tempo_all, 
                         mapping = aes(x = duration, 
                                       y = tempo, 
                                       color = tempo)) + 
  geom_point() + 
  labs(x = "Duration", y = "Tempo") + 
  ggtitle("Tempo Over Duration") +
  theme(plot.title = element_text(hjust = 0.5))
dur_tempo_plot + scale_color_gradient()
```
### Analyze correlation and model for duration and tempo
The correlation between duration and tempo was 0.003. A value closer to 0 suggests a weak relationship between the variables. A low correlation (-0.2 < x < 0.2) probably suggests that much of variation of the response variable, tempo, is unexplained by the predictor, duration, in which case, we should probably look for better explanatory variables.

Just in case though, we decide to model duration vs tempo and create a linear regression model. The null hypothesis is that the coefficients associated with the variables is equal to zero. The alternate hypothesis is that the coefficients are not equal to zero. There exists a relationship between the independent variable in question (duration in this case) and the dependent variable (tempo). As the p-value is above 0.05, this means that the null hypothesis is accepted and there does not appear to be a relationship between duration and average tempo. The linear model is not statistically significant as the p-value of 0.173 is higher than the significance level 0.05. 

Finally, the R-squared value is extremely low at 7.6e-06 meaning almost none of the variation in the average tempo has been explained by this model.
```{r}
# bind all columns together from original data with tempo outliers removed into a df
all <- cbind(tempo_no_outliers, tempo_years, df_tempo_all$duration)
df_all <- as.data.frame(all)
# rename columns
names(df_all) <- c("tempo", "year", "duration")
df_all

#count how many outliers as NAs
na_df_all_count <- df_all %>% 
  summarise(na_count = sum(is.na(tempo)))

# all variables df with outlier/NAs removed
df_all_df <- df_all %>% filter(!is.na(tempo))

#correlation between tempo and duration
cor(df_all_df$tempo, df_all_df$duration) # no correlation
# model tempo and duration
tempo_dur_model <- lm(tempo ~ duration, data = df_all_df)
summary(tempo_dur_model)
```
